Rusty Linux: Advances in Rust for Linux Kernel Development

### 2024-10-24 slides.tex
Removed a comma from my name
Minor updates to slides

%\useoutertheme{shadow}
%\useinnertheme{rounded}

@@

\author[]
{Shane K. Panter\inst{1} \and Nasir, Eisty\inst{2}}

@@

@@

\subsection{RQ1: Existing Approaches}
\begin{frame}
  \frametitle{RQ1}
  \begin{table}
    \caption{Approaches and Methodologies for Rust in the Kernel}
    \begin{tabular}{||l|l|l||}
      \hline
      Approach & Papers & Operating System in Rust\\
      \hline\hline
      Monolithic   & 4 & Linux Kernel v6.1+\\
      Micro-kernel & 5 & Atmosphere, Redox, Redleaf\\
      Embedded     & 2 & Tock, Hubris, Drone, Bern, HarSaRK \\
      Unikernel & 4 & RustyHermit, Theseus \\
      Exokernel & 1 & W-Kernel \\
      \hline
    \end{tabular}
    \label{tab:RQ1}
  \end{table}

@@

\subsection{RQ2: Performance Implications}
\begin{frame}
  \frametitle{RQ2}
  \begin{table}
    \caption{Performance Implications of Rust in the Kernel}
    \begin{tabular}{||l|l|l||}
      \hline
      No. & Implication & Studies that Reported the challenge\\
      \hline\hline
      1 & Performance & 3\\
      2 & Throughput & 1\\
      3 & Latency & 1 \\
      \hline
    \end{tabular}
    \label{tab:RQ2}
  \end{table}
\end{frame}

\subsection{RQ3: Challenges and Limitations}
\begin{frame}
  \frametitle{RQ3: Challenges and Limitations}
  \begin{itemize}
  \item<1-> Binary Size - Rust can produce larger binaries
  \item<2-> Missing Features - Rust still evolving and adding features
  \item<3-> Soundness - How to deal with raw memory
  \item<4-> Panics - What happens when things go wrong?
  \item<5-> C Interop - Specific to mixed language kernels
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{RQ4: Lessons Learned}
  \begin{itemize}
  \item<1-> Impossible to use 100$\%$ rust
  \item<2-> Rust is not as expressive as other formal verification techniques
  \item<3-> Ownership root - An OS provides memory to rust so if the OS is itself written in rust
    who is the root owner?
  \end{itemize}

@@

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
  \item<1-> We are still in the early stages of figuring out who to do kernel dev in Rust
  \item<2-> High potential for enhanced security and stability
  \item<3-> Need to address integration issues (FFI)
  \item<4-> Need to expand the body of empirical evidence on Rust's impact!
  \end{itemize}

>>>>

%\useoutertheme{split}
%\useoutertheme{shadow}
%\useoutertheme{sidebar}
%\useinnertheme{rounded}
%\useinnertheme{inmargin}

@@

\author[]
{Shane K. Panter\inst{1} \and Nasir Eisty\inst{2}}

@@

\section{Why Rust}
\begin{frame}
  \frametitle{Why Rust}
  \begin{itemize}
  \item<1-> Low-level \textbf{control} like C and C++
  \item<2-> Strong \textbf{safety} guarantees
  \item<3-> \textbf{Modern}, functional paradigms
  \item<4-> Industrial development and backing
  \item<5-> No garbage collector needed! All checks are performed at compile time
  \end{itemize}
\end{frame}

@@

\subsection{RQ1: Existing Approaches}
\begin{frame}
  \frametitle{RQ1}
  \begin{table}
    \caption{Approaches and Methodologies for Rust in the Kernel}
    \begin{tabular}{||l|l|l||}
      \hline
      Approach & Papers & Operating System in Rust\\
      \hline\hline
      Monolithic   & 4 & Linux Kernel v6.1+\\
      Micro-kernel & 5 & Atmosphere, Redox, Redleaf\\
      Embedded     & 2 & Tock, Hubris, Drone, Bern, HarSaRK \\
      Unikernel & 4 & RustyHermit, Theseus \\
       \hline
    \end{tabular}
    \label{tab:RQ1}
  \end{table}

@@

\subsection{RQ2: Performance Implications}
\begin{frame}
  \frametitle{RQ2}
  \begin{table}
    \caption{Performance Implications of Rust in the Kernel}
    \begin{tabular}{||l|l|l||}
      \hline
      No. & Implication & Studies that Reported the challenge\\
      \hline\hline
      1 & Performance & 3\\
      2 & Throughput & 1\\
      3 & Latency & 1 \\
      \hline
    \end{tabular}
    \label{tab:RQ2}
  \end{table}

  \begin{itemize}
  \item Performance issues - Caused by the safe $->$ unsafe transition layer
  \item Throughput issues - Caused by immature and or missing bindings within the FFI layer
  \item Latency issues - Caused by the interrupt layer written in Rust
  \end{itemize}

\subsection{RQ3: Challenges and Limitations}
\begin{frame}
  \frametitle{RQ3: Challenges and Limitations}
  \begin{itemize}
  \item<1-> Binary Size
    \begin{itemize}
    \item Rust can produce larger binaries
    \item The same issue that C++ templates have!
    \end{itemize}
  \item<2-> Missing Features
    \begin{itemize}
    \item Rust still evolving and adding features
    \item Makes it difficult to integrate into the classroom due to the rapid evolution
    \end{itemize}
  \item<3-> Soundness - How to deal with raw memory?
  \item<4-> Panics - What happens when things go wrong?
  \item<5-> C Interop - Specific to mixed language kernels
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{RQ4: Lessons Learned}
  \begin{itemize}
  \item<1-> Impossible to use 100$\%$ rust
  \item<2-> Rust is not as expressive as other formal verification techniques
  \item<3-> Ownership root - An OS provides memory to rust so if the OS is itself written in rust
    who is the root owner?
    \begin{itemize}
      \item Research is needed for hardware support (CHERI)
  \end{itemize}

@@

\begin{frame}
  \frametitle{Conclusion}
  \begin{itemize}
  \item<1-> We are still in the early stages of figuring out who to do kernel dev in Rust
  \item<2-> High potential for enhanced security and stability
  \item<3-> Need to address integration issues (FFI)
  \item<4-> Need to expand the body of empirical evidence on Rust's impact! (Or more generally low
    level memory safe languages)
  \end{itemize}
